A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-09 paper.tex

Figures \ref{fig:polypower} and \ref{fig:heterpower} present the power
curves of various tests plotted against the effect size in the
residuals. The effect size was computed at a 5\% significance level and
plotted on a natural logarithmic scale. To facilitate visual calibration
of effect size values with the corresponding diagnostic plots, a
sequence of example residual plots with increasing effect sizes is
provided at the bottom of these figures. These plots serve as a visual
aid to help readers understand how different effect size values
translate to changes in the diagnostic plots.

displays the estimated power of a visual test, computed on evaluations
of the lineups generated by the non-linearity model with uniform fitted
values. Power is plotted against the natural logarithm of the effect
size \(log_e(E)\), computed at a 5\% significance level. The small plots
at the bottom of this figure shows a sequence of example residual plots
of increasing effect size. These are for the readers to compare what a
residual plot might look like at particular effect sizes.

Figure \ref{fig:heterpower} is similar to Figure \ref{fig:polypower},
but shows corresponding information on lineups produced by the
heteroskedasticity model. In this scenario, the visual test is compared
to an approximate test - Breusch-Pagan test, and two other inappropriate
tests - RESET test and Shapiro-Wilk test.

For non-linearity patterns, the power curve of RESET test climbs
aggressively from 7\% to 84\% as \(log_e(E)\) increases from 0 to 2.5,
while power of other tests respond inactively to the change of effect,
showing that RESET test is much more sensitive to weak non-linear
structure. Meanwhile, no noticeable residual departures can be spotted
from the example residual plots.

In terms of heteroskedasticity patterns, the power of Breusch-Pagan test
is constantly greater than the power of visual test. At
\(log_e(E) \approx 2.5\), where the power curve of the visual test
remains at a low level, the Breusch-Pagan test has around 50\% chance of
rejecting \(H_0\). Similarly, the visual feature is nearly unobservable
from the example residual plots at this level of effect size.

The power of visual test arises steadily as \(log_e(E)\) increases from
3 to 5 for both non-linearity patterns and heteroskedasticity patterns,
suggesting that the effect starts to make significant impact on the
degree of the presence of the designed visual features. This can also be
observed from the example residual plots that when \(log_e(E) = 3.5\), a
weak ``S-shape'' and a weak ``right-triangle'' shape are presented in
Figure \ref{fig:polypower} and Figure \ref{fig:heterpower} respectively.
The visual pattern becomes clearer as \(log_e(E)\) increases. At
\(log_e(E) = 5\), the power of visual tests for both patterns reaches
almost 100\%.

Power curves of inappropriate tests show improvement as the effect
increases but at a lower rate than the visual test in both scenarios.
This coincides the point made by \citet{cook1982residuals} that
residual-based tests for a specific type of model defect may be
sensitive to other types of model defects. The power curve of RESET test
remains at around 5\% in Figure \ref{fig:heterpower} since there are no
non-linear terms leave out in the heteroskedasticity model and \(H_0\)
of the test is always satisfied.

Overall, the power comparison suggests that conventional tests differs
significantly from visual tests in two regression diagnostics scenarios
designed by us. Visual test have much higher tolerance of the residual
departures than the conventional test. Since fail to reject \(H_0\) in a
visual test literally means that there are no obvious visual discoveries
found in the residual plot, analysts and the general public as the
consumers of the output may not be convinced of the existence of
significant residual departures in spite of the rejection of \(H_0\)
given by the conventional test. Even if the rejection is accepted, the
model violation may be considered as impactless due to the fact that
they are not clearly visible. Besides, the sensitivity of the
conventional test to weak residual departures could also distract and
discourage analysts from finding simple but good linear approximation to
the data. The rejection of \(H_0\) because of human acceptable and
negligible residual departures is not practically meaningful and useful
to analysts and decision makers. In contrast, if the strict correctness
of the model assumption is of particular interest, conducting a
conventional test is still the recommended choice based on the findings.

@@

\caption{Comparison of power between different tests for non-linear patterns (uniform fitted values only). The power curves are estimated using logistic regression, and the horizontal lines of dots represent decisions made by visual tests, with 0, 1 indicating non-reject and reject, respectively. The visual test has multiple power curves estimated from bootstrap samples. The row of scatterplots plots at the bottom are examples of residual plots corresponding to the specific effect sizes marked by vertical lines in the main plot.}\label{fig:polypower}
\end{figure}

@@

\caption{Comparison of power between different tests for heteroskedasticity patterns (uniform fitted values only). Main plot shows the power curves, with dots indicating human evaluations of lineups. Surrounding lines of the visual test show the estimated power of 500 bootstrap samples. Small row of plots shows typical residual plots corresponding to specific effect sizes, marked by vertical lines in main plot. Where would you draw the line of too much heteroskedasticity in the residuals? For the BP test this is around log effect size 2.5, but for the visual test it is around 3.5.}\label{fig:heterpower}
\end{figure}

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

>>>>

@@

\caption{Comparison of power between different tests for non-linear patterns (uniform fitted values only). The power curves are estimated using logistic regression, and the horizontal lines of dots represent non-detect and detect results from human observers for each lineup. The visual test has multiple power curves estimated from bootstrap samples. The row of scatterplots plots at the bottom are examples of residual plots corresponding to the specific effect sizes marked by vertical lines in the main plot.}\label{fig:polypower}
\end{figure}

@@

\caption{Comparison of power between different tests for heteroskedasticity patterns (uniform fitted values only). Main plot shows the power curves, with dots indicating non-detect and detect in human evaluations of lineups. The multiple lines for the visual test arise from estimating the power on many bootstrap samples. The row of scatterplots plots at the bottom are examples of residual plots corresponding to the specific effect sizes marked by vertical lines in the main plot.}\label{fig:heterpower}
\end{figure}

Figures \ref{fig:polypower} and \ref{fig:heterpower} present the power
curves of various tests plotted against the effect size in the
residuals, for non-linearity and heteroskedasticity, respectively. In
each case the power of visual test is re-calculated for multiple
bootstrap samples leading to the many (orange) curves. The effect size
was computed at a 5\% significance level and plotted on a natural
logarithmic scale. To facilitate visual calibration of effect size
values with the corresponding diagnostic plots, a sequence of example
residual plots with increasing effect sizes is provided at the bottom of
these figures. These plots serve as a visual aid to help readers
understand how different effect size values translate to changes in the
diagnostic plots. The horizontal lines of dots at 0 and 1 represent the
detects or non-detects made by human observers for each lineup.

Figures \ref{fig:polypower} compares the power for the different tests
for non-linear structure in the residuals. The test with the uniformly
higher power is the RESET test, one that specifically tests for
non-linearity. The power curves for the visual test are effectively a
shift right from that of the RESET test. This means that the RESET test
will reject a lower effect size (less structure) than the visual test,
but otherwise the performance will be similar. In other words, the RESET
test is more sensitive than the visual test. This is not necessarily a
good feature, for the purposes of diagnosing model defects. If one scans
the residual plot examples at the bottom, we might argue that the
non-linearity is not sufficiently problematic until an effect size of
around 3 or 3.5. The RESET test would reject closer to an effect size of
2, but the visual test would reject closer 3.25. As expected the
Breusch-Pagan and Shapiro-Wilk tests have much lower power - they are
not designed to detect non-linearity.

For the heteroskedasticity pattern, the power of Breusch-Pagan test,
designed for detecting heteroskedasticity, is uniformly higher than the
other tests. The visual test power curve is a right shift. This shows a
similar story to the power curves for non-linearity pattern - the
conventional test is more sensitive than the visual test. From the
example residual plots at the bottom we might argue that the
heteroskedasticity becomes noticeably visible around an effect size of 3
or 3.5. However the Breusch-Pagan test would reject at around effect
size 2.5. Interestingly, the power curve for the Shapiro-Wilk test (for
non-normality) is only sighly different to that of the visual test,
suggesting that it performs reasonably for detecting heteroskedasticity,
too. The power curve for the Breusch-Pagan test suggests it is not
useful for detecting heteroskeadsticity, as expected.

Overall, the results show that the conventional tests are more sensitive
than the visual test. The conventional tests do have higher power for
the patterns they are designed to detect, and generaly unable to detect
other patterns. The visual test doesn't require specifiying the pattern
ahead of time, relying purely on whether the observed residual plot is
detectably different from ``good'' residual plots. They will perform
equally well regardless of the type of model defect. This aligns with
the advice of experts on residual analysis, who consider residual plot
analysis to be an indispensable tool for diagnosing model problems. What
we gain from using a visual test for this purpose is the removal of
subjective argument about whether a pattern is visible or not. The
lineup protocol provides the calibration of detecting patterns, that if
the pattern in the data plot cannot be distinguished from patterns in
good residual plots, then no discernible problem with the model exists.

@@

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}
